The investment strategy based on behavior of artificial earthworm for use in algotrading

### 2014-05-09 Article/Article20140406Granada/LatexArt/article_main.tex
italic

Artificial earthworm executes the following algorithm.\\
The strategy is built on the principle of classical machine learning. Before starting to run the algorithm, on data located outside the environment that investor knows (like on the current market segments --- out-of-sample), it is necessary to ,,teach''  the values of parameters. This principle is presented in Figure \ref{fig:fig2}, which shows examples of learning periods and following them testing periods. Data from testing period was not used during learning.\\
\begin{figure}[ht]
\centering
%\includegraphics[width = 0.6\textwidth]{figures/rys2.png}
\caption{An example of the arrangement of learning and testing periods of strategy parameters in time series.}
\label{fig:fig2}
\end{figure}
\FloatBarrier
\vspace{-1em}
This repeatable algorithm, located in both periods - learning and testing, has the following structure. In learning period the optimal value for five parameters $p1$, $p2$, ..., $p5$ was determined. In test period these designated values of parameters were used to perform the algorithm in the new environment. In each $i-th$ step (candle) of time series following activities were performed:
\begin{itemize}
\item For each $i = p1, p1 +1, p1 +2, ..., I$ and for $j = 1, ..., p1$ check whether the candle located at a distance i-j backwards from the current candle is:\\
green when $O_{i-j} <C_{i-j}$, \\
red when $O_{i-j}> C_{i-j}$, \\
where $p1$ --- is the first parameter which is the length of the artificial worm - the number of candles of different colors that make up the ,,body'' of the worm; \\
$O_{i-j}$ --- is the opening value of (i-j)-th candle;\\ 
$C_{i-j}$ --- is the closing value of (i-j)-th candle; \\
$I$ --- is the length of the test window, after which it is necessary to change the parameters values to the new, learned in the machine learning phase;\\
\item In each $i-th$ step, the loop $j = 1, ..., p1$ is performed in which a number of green $Lg$, red $Lr$ candles is calculated so the sum of $Lg$ and $Lr$ is $p1$ (to avoid the situation when $O_i = C_i$, it is assigned arbitrarily to one of the candles sets, for example green);
\item A condition for the existence of a Rg recommendation to open a long position (or to continue it if it is open) is checked. This condition is arbitrarily adopted (one can of course consider different one): 
\begin{equation}
\text{If } Lg> p1/2-p2 \text{ then } Rg=1
\end{equation}

\end{itemize}
If this condition is fulfilled, then depending on the current position status PositionState (where PositionState is a variable which takes values {long, short, 0}) at the opening of the next candle, a position is opened or continued as follows: 
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState =long \text{ then PositionState =long;}
\end{equation}
(costless continuation of opened position);
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState = short \text{ then PositionState=long;}
\end{equation}
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState =0 \text{ then PositionState =long;}
\end{equation}

Similarly check the condition of recommendations Rr to open or continue short position:
\begin{equation}
\text{If } Lr>p1/2 – p2 \text {then Rr=1.}
\end{equation}
If this condition is fulfilled, then depending on what the current position status PositionState is, at the opening of the next candle, position is opened or continued as follows: \\
\begin{equation}
\text{If } Rr=1 \text{ and } PositionState =long \text{ then PositionState =short;}
\end{equation}
\begin{equation}
\text{If } Rr==1 \text{ and } PositionState ==short \text{ then PositionState=short;}
\end{equation}
(costless continuing of opened position)
\begin{equation}
\text{If } Rr==1 \text{ and }  PositionState ==0 \text{ then  PositionState =short;}
\end{equation}

In formulas above there is a parameter $p2$. It is the second parameter of the strategy which is an arbitrarily separated part of the worm, which depending on size causes the direction of investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of $p2$ leads to a higher frequency of these changes. The optimum value of $p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the spread = 0, when opened long position is continued; \\
and spread = cost, when long position was opened after closing or opening a short position in case when there was no open position;\\
cost is a variable that depends on the particular brokerage platform that was used for testing the strategy; in this case, for the currency pair EURUSD $cost = 0.00016$ was assumed and for the currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the $spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or opening a long position in case when there was no open position.
\end{itemize}

@@

\noindent Presented strategy was tested if:
\begin{itemize}
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a long position presented condition did not occur:\\
\begin{equation}
\text{If } L_i-O_i<SL \text{ then } Z_i=-SL
\end{equation}
where Li --– the lowest rate value in the $i-th$ candle; $SL$ --– absolute value Stop Loss appearing as $p5$ parameter;
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a short position presented condition did not occur:
\begin{equation}
\text{If } O_i-H_i<SL \text{ then } Z_i=-SL
\end{equation}
where Hi --– the highest rate value in the $i-th$ candle.
\end{itemize}
This way $p5$ parameter was used for securing in strategy the strong local changes in exchange rate within a single candle.\\
The meaning of $p5$ parameter to the behavior of the worm and the occurrence of the closing position is explained in Figure \ref{fig:fig5}.
\begin{figure}[h!]
\centering
\includegraphics[width = 0.5\textwidth]{figures/rys5.png}
\caption{Method of calculating the profit (here: loss) in the case of necessity to use $p5$ parameter actuating mechanism $SL$ for long positions.}
\label{fig:fig5}
\end{figure}
\FloatBarrier

\begin{itemize}
\item To determine the current quality of the strategy and risk management summation of current gains/losses (the accumulation) is performed
\begin{equation}
Zc_i = sum ( Z_i ) \text{ from } i = 1 \text{ to the current i value;}
\end{equation}
\item There is a possible local subsidence $Zc_i$ capital by introducing two additional parameters to manage the risk.\\
$p3$ --- the number of steps back on the curve of the cumulative profit for the determination of $Zc_i$ --- $p3$ for comparison with the present value of $Zc_i$;
$p4$ --- maximum size of the capital subsidence (eg for the currency pair EURUSD measured in pips), which if exceeded will trigger the closure position:\\
\begin{equation}
\text{If } ( Zc_i - p3 - Z_i ) <- p4 \text{ and } PositionState \neq 0 \text{ then } PositionState = 0
\end{equation} 
where PositionState --- is the type of open position (long or short) and PositionStates equal 0 means resignations from the opening position to the beginning of the next candle.
\end{itemize}
So, the above-described algorithm uses 5 parameters, as many as 3 of which are devoted to risk management of the strategy.\\
%%PBStart
The complexity of the algorithm is illustrated in Figure \ref{fig:fig7}, which shows the individual steps and decisions made in the iteration on a single candle in the form of a decision tree. This iteration was carried out both in the search for optimal parameters $p1, ..., p5$ and in the process they use in the testing phase for data not used in learning period.\\
\begin{figure}[h!]
\centering
\includegraphics[width = \textwidth]{figures/rys7.png}
\caption{Decision tree for the states and decisions in one step (candle) in the learning phase and testing.}
\label{fig:fig7}
\end{figure}
\FloatBarrier
\vspace{-1em}
The figure shows in a tree form further decisions taken at the beginning of the candle and then reactions to any unacceptable change in course. The first examines the state of the position taken over from the closure of the previous candle. The next step is to work out decisions / recommendations as to the direction of the opening from the set {long, short, none}. After opening position is observed return of the open position. It can be positive or negative. If it is positive, it does not take up any additional decisions. If it is negative, it checks if it is not achieved the level of $SL$. In the case of a negative return of the items, except for testing, $SL$ is not exceeded within a candle tested is also a condition resulting from the use of parameters ($p3$, $p4$). Tree is a state of complete items transmitted at the beginning of the next candle.

>>>>

Artificial earthworm executes the following algorithm.\\
The strategy is built on the principle of classical machine learning. Before starting to run the algorithm, on data located outside the environment that investor knows (like on the current market segments --- out-of-sample), it is necessary to ,,teach''  the values of parameters. This principle is presented in Figure \ref{fig:fig2}, which shows examples of learning periods and following them testing periods. Data from testing period was not used during learning.\\
\begin{figure}[ht]
\centering
%\includegraphics[width = 0.6\textwidth]{figures/rys2.png}
\caption{An example of the arrangement of learning and testing periods of strategy parameters in time series.}
\label{fig:fig2}
\end{figure}
\FloatBarrier
\vspace{-1em}
This repeatable algorithm, located in both periods - learning and testing, has the following structure. In learning period the optimal value for five parameters $p1$, $p2$, ..., $p5$ was determined. In test period these designated values of parameters were used to perform the algorithm in the new environment. In each $i-th$ step (candle) of time series following activities were performed:
\begin{itemize}
\item For each $i = p1, p1 +1, p1 +2, ..., I$ and for $j = 1, ..., p1$ check whether the candle located at a distance i-j backwards from the current candle is:\\
green when $O_{i-j} <C_{i-j}$, \\
red when $O_{i-j}> C_{i-j}$, \\
where $p1$ --- is the first parameter which is the length of the artificial worm - the number of candles of different colors that make up the ,,body'' of the worm; \\
$O_{i-j}$ --- is the opening value of (i-j)-th candle;\\ 
$C_{i-j}$ --- is the closing value of (i-j)-th candle; \\
$I$ --- is the length of the test window, after which it is necessary to change the parameters values to the new, learned in the machine learning phase;\\
\item In each $i-th$ step, the loop $j = 1, ..., p1$ is performed in which a number of green $Lg$, red $Lr$ candles is calculated so the sum of $Lg$ and $Lr$ is $p1$ (to avoid the situation when $O_i = C_i$, it is assigned arbitrarily to one of the candles sets, for example green);
\item A condition for the existence of a $Rg$ recommendation to open a long position (or to continue it if it is open) is checked. This condition is arbitrarily adopted (one can of course consider different one): 
\begin{equation}
\text{If } Lg> p1/2-p2 \text{ then } Rg=1
\end{equation}

\end{itemize}
If this condition is fulfilled, then depending on the current position status $PositionState$ (where $PositionState$ is a variable which takes values {long, short, 0}) at the opening of the next candle, a position is opened or continued as follows: 
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState =long \text{ then }PositionState =long;
\end{equation}
(costless continuation of opened position);
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState = short \text{ then } PositionState=long;
\end{equation}
\begin{equation}
\text{If } Rg=1 \text{ and } PositionState =0 \text{ then } PositionState =long;
\end{equation}

Similarly check the condition of recommendations Rr to open or continue short position:
\begin{equation}
\text{If } Lr>p1/2 – p2 \text {then Rr=1.}
\end{equation}
If this condition is fulfilled, then depending on what the current position status PositionState is, at the opening of the next candle, position is opened or continued as follows: \\
\begin{equation}
\text{If } Rr=1 \text{ and } PositionState =long \text{ then } PositionState =short;
\end{equation}
\begin{equation}
\text{If } Rr==1 \text{ and } PositionState ==short \text{ then } PositionState=short;
\end{equation}
(costless continuing of opened position)
\begin{equation}
\text{If } Rr==1 \text{ and }  PositionState ==0 \text{ then }  PositionState =short;
\end{equation}

In formulas above there is a parameter $p2$. It is the second parameter of the strategy which is an arbitrarily separated part of the worm, which depending on size causes the direction of investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of $p2$ leads to a higher frequency of these changes. The optimum value of $p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the $spread = 0$, when opened long position is continued; \\
and $spread = cost$, when long position was opened after closing or opening a short position in case when there was no open position;\\
$cost$ is a variable that depends on the particular brokerage platform that was used for testing the strategy; in this case, for the currency pair EURUSD $cost = 0.00016$ was assumed and for the currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the $spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or opening a long position in case when there was no open position.
\end{itemize}

@@

\noindent Presented strategy was tested if:
\begin{itemize}
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a long position presented condition did not occur:\\
\begin{equation}
\text{If } L_i-O_i<SL \text{ then } Z_i=-SL
\end{equation}
where $L_i$ --– the lowest rate value in the $i-th$ candle; $SL$ --– absolute value Stop Loss appearing as $p5$ parameter;
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a short position presented condition did not occur:
\begin{equation}
\text{If } O_i-H_i<SL \text{ then } Z_i=-SL
\end{equation}
where Hi --– the highest rate value in the $i-th$ candle.
\end{itemize}
This way $p5$ parameter was used for securing in strategy the strong local changes in exchange rate within a single candle.\\
The meaning of $p5$ parameter to the behavior of the worm and the occurrence of the closing position is explained in Figure \ref{fig:fig5}.
\begin{figure}[h!]
\centering
\includegraphics[width = 0.5\textwidth]{figures/rys5.png}
\caption{Method of calculating the profit (here: loss) in the case of necessity to use $p5$ parameter actuating mechanism $SL$ for long positions.}
\label{fig:fig5}
\end{figure}
\FloatBarrier

\begin{itemize}
\item To determine the current quality of the strategy and risk management summation of current gains/losses (the accumulation) is performed
\begin{equation}
Zc_i = sum ( Z_i ) \text{ from } i = 1 \text{ to the current i value;}
\end{equation}
\item There is a possible local subsidence $Zc_i$ capital by introducing two additional parameters to manage the risk.\\
$p3$ --- the number of steps back on the curve of the cumulative profit for the determination of $Zc_i$ --- $p3$ for comparison with the present value of $Zc_i$;
$p4$ --- maximum value measured in pips of the capital drawdown (eg for the currency pair EURUSD), which if exceeded will trigger the closure position:\\
\begin{equation}
\text{If } ( Zc_i - p3 - Z_i ) <- p4 \text{ and } PositionState \neq 0 \text{ then } PositionState = 0
\end{equation} 
where $PositionState$ --- is the type of open position (long or short) and $PositionState$ equal 0 means resignations from the opening position to the beginning of the next candle.
\end{itemize}
So, the above-described algorithm uses 5 parameters, as many as 3 of which are devoted to risk management of the strategy.\\
%%PBStart
The complexity of the algorithm is illustrated in Figure \ref{fig:fig7}, which shows the individual steps and decisions made in the iteration on a single candle in the form of a decision tree. This iteration was carried out both in the search for optimal parameters $p1, ..., p5$ and in the process they use in the testing phase for data not used in learning period.\\
\begin{figure}[h!]
\centering
\includegraphics[width = \textwidth]{figures/rys7.png}
\caption{Decision tree for the states and decisions in one step (candle) in the learning phase and testing.}
\label{fig:fig7}
\end{figure}
\FloatBarrier
\vspace{-1em}
The figure shows in a tree form further decisions taken at the beginning of the candle and then reactions to any unacceptable change in course. The first examines the state of the position taken over from the closure of the previous candle. The next step is to work out decisions / recommendations as to the direction of the opening from the set {long, short, none}. After opening position is observed return of the open position. It can be positive or negative. If it is positive, it does not take up any additional decisions. If it is negative, it checks if it is not achieved the level of $SL$. In the case of a negative return of the items, except for testing, $SL$ is not exceeded within a candle tested is also a condition resulting from the use of parameters ($p3$, $p4$). Tree is a state of complete items transmitted at the beginning of the next candle.
